---
title: "Rough Draft: An analysis of individual stochastic terrorists in the Global Terrorism Databasse"
subtitle: "Application of Advanced Frameworks for Data Imputation in the Global Terrorism Database"
author: "James Hunter"
bibliography: ds_jjh_bib.bib
csl: american-political-science-association.csl
abstract: >
 This paper explores the ability of unsupervised machine learning algorithms to perform data imputation against the Global Terrorism Database (GTD), a large, complex dataset that has been carefully curated to contain a comprehensive listing of terrorist attacks from 1970 to 2021. The research explores the ability to populate missing fields by contextual processing of full-text fields, with a particular emphasis on synthesizing data to perform a longitudinal analysis of group size on attack tactics and outcomes.
papersize: letter

margin-left: 1in
margin-right: 1in
margin-top: 1in
margin-bottom: 1in
cap-location: top

format: 
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    linestretch: 2
    
execute:
  echo: false
  warning: false
  message: false
---

\clearpage

\raggedright

\setlength{\parindent}{20pt}

```{r notes, include=FALSE}
#Your rough draft should be about 10 pages and should represent good progress towards your final submission. You should have 2-3 pages of introduction that states a clear research question, followed by 4-5 pages of factual and theoretical development based on previous literature. That development section should set up a statement of a hypothesis. After that, use 3-4 pages to describe a strategy for testing that hypothesis using quantitative methods, based on data that you have access to and have managed/cleaned/prepared. Be as specific as you possibly can with this methodology. You must have results to share at this point, demonstrating that you have command of your data and can generate insights from it.
```

```{r package_install, include=FALSE}
remotes::install_github("https://github.com/r-forge/countreg", subdir="pkg")
install.packages("topmodels", repos = "https://R-Forge.R-project.org")
```

```{r setup, include=FALSE}
## Libraries
library(lmtest)   # LM comparison
library(sandwich) # vcovHC lives here
library(ggplot2)  # Plots
library(stargazer)# Multiple RSEs
library(knitr)    # Nice table
library(AER)      # Canned data for testing
library(dplyr)    # Slice and dice data
library(readr)    # Load CSVs fancier
library(lubridate)# Date/Time formatting
library(tidyr)
library(purrr)    # Additional dataframe translations
library(knitr)    # Basic table formatting
library(kableExtra)
library(webshot2) # export images for knitr
library(magick)   # Improve quality of exported tables
library(ggExtra)  # More cool plots
library(mice)     # R built-in logic to handle MICE for nperps
library(Metrics)  # For RMSE and MAE calculations
library(caret)    # Confusion matrix
library(broom)    # Forest Plot
library(broom.mixed) # additional tidy() functions
library(pscl)     # ZINB regression
library(remotes)  # Grab repos not on CRAN FORGE
library(gridExtra)# side-by-side models
library(patchwork)# another attempt at side-by-side models
library(countreg) # rootogram
library(topmodels)# another rootogram implementation
```

```{r data_load, include=FALSE}
#| cache: true
## Data Load (Get a cuppa coffee...)
gtd_full <- read.csv('.\\Data\\globalterrorismdb_0522dist.csv')
acled_full <- read.csv('.\\Data\\ACLED_Data_2026-02-08.csv')
```

```{r data_shape, include=FALSE}
#| cache: true
# Replace GTD missing value codes (-99) with NA
gtd_full <- gtd_full %>%
  mutate(
    nperps = ifelse(nperps == -99 | nperps == "Unknown", NA, nperps),
    nkill = ifelse(nkill == -99 | nkill == "Unknown", NA, nkill),
    nkillus = ifelse(nkillus == -99 | nkillus == "Unknown", NA, nkillus),
    nkillter = ifelse(nkillter == -99 | nkillter == "Unknown", NA, nkillter),
    nwound = ifelse(nwound == -99 | nwound == "Unknown", NA, nwound),
    gname = ifelse(gname == -99 | gname == "Unknown", NA, gname),
    gname2 = ifelse(gname2 == -99 | gname2 == "Unknown", NA, gname2),
    gname3 = ifelse(gname3 == -99 | gname3 == "Unknown", NA, gname3),
    attacktype1 = ifelse(attacktype1 == -99 | attacktype1 == "Unknown", NA, attacktype1),
    weaptype1 = ifelse(weaptype1 == -99 | weaptype1 == "Unknown", NA, weaptype1),
    targtype1 = ifelse(targtype1 == -99 | targtype1 == "Unknown", NA, targtype1),
    suicide = ifelse(suicide == -99 | suicide == "Unknown", NA, suicide),
    region = ifelse(region == -99 | region == "Unknown" | region == "", NA, region),
    motive = ifelse(motive == -99 | motive == "motive" | motive == "", NA, motive)
  )

## Shape the GTD and to the same time bounds as the ACLED (2000-2021)
gtd_clean <- gtd_full %>%
  # Remove rows where month or day is 0 (unknown) so we can make valid dates
  filter(imonth > 0, iday > 0) %>%
  mutate(
    event_date = make_date(iyear, imonth, iday),
    lat_round = round(latitude, 1),
    long_round = round(longitude, 1)
  ) %>%
  filter(event_date >= as.Date("2000-01-01") & event_date <= as.Date("2021-01-01")) %>%
  filter(is.na(nperps) | nperps < 50)

acled_clean <- acled_full %>%
  mutate(
    event_date = as.Date(event_date),
    lat_round = round(latitude, 1),
    long_round = round(longitude, 1)
  )

# Identify Shared Rows and consolidate
shared_events <- inner_join(
  acled_clean, 
  gtd_clean, 
  by = c("event_date", "lat_round", "long_round"),
  suffix = c("_acled", "_gtd")
)

print(paste("Number of shared events found:", nrow(shared_events)))
head(shared_events)
#rm(gtd_full, acled_full)
nperps_one <- gtd_clean %>%
  filter(nperps == 1) %>%
  nrow()

nperps_more <- gtd_clean %>%
  filter(nperps >= 1) %>%
  nrow()
```

```{r gtd_missingness, include=FALSE}
#| cache: true
gtd_target_cols <- c("gname", "nperps", "nkill", "nwound", "motive")
gtd_target_col_labels <- c("Group Name", "Number of Perpetrators", "Number Killed", "Number Wounded", "Motive")
  
gtd_missing_data_summary <- gtd_clean %>%
  dplyr::select(all_of(gtd_target_cols)) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  # Reshape data from wide to long format for plotting
  pivot_longer(cols = everything(), 
               names_to = "Column", 
               values_to = "Missing_Pct") %>%
  mutate(Column = reorder(Column, +Missing_Pct))
```

```{r acled_missingness, include=FALSE}
#| cache: true
acled_target_cols <- c("actor1", "actor2", "inter1", "inter2")
aclet_target_col_labels <- c("Primary Actor", "Secondary Actor", "Interactor", "Secondary Interactor")

acled_missingness_summary <- acled_clean %>%
  dplyr::select(all_of(acled_target_cols)) %>%
  summarise(across(everything(), function(x) {
    # Calculate percentage where ANY of these conditions are true:
    mean(
      is.na(x) |                        # Standard NA
      x == "" |                         # Empty string
      grepl("^\\s*$", x) |              # String with only whitespace
      grepl("Unidentified", x, ignore.case = TRUE) # Contains "Unidentified"
    ) * 100
  })) %>%
  pivot_longer(everything(), names_to = "Column", values_to = "Missing_Pct")
```

```{r mice_nperps, include=FALSE}
#| cache: true
# Generate full predictor matrix, divide nperps into categorical variable
gtd_clean <- gtd_clean %>%
  mutate(nperps = if_else(suicide == 1 & is.na(nperps), 1, nperps)) %>%
  mutate(
    nperps_cat = case_when(
      nperps == 1 ~ "Lone Wolf",
      nperps > 1 & nperps <= 5 ~ "Small Cell",
      nperps > 5 ~ "Large Group",
      # Leave as NA where the original nperps is NA
      TRUE ~ NA_character_ 
    )
  ) %>%
  # MICE requires categorical targets to be explicitly defined as factors
  mutate(nperps_cat = as.factor(nperps_cat))
# Select only the columns needed for the imputation model to save memory/time
impute_data <- gtd_clean %>%
  dplyr::select(eventid, nperps, nperps_cat, nkill, nwound, attacktype1, weaptype1, targtype1, suicide, region, country, iyear, country_txt, region_txt, provstate)

# Convert categorical variables to factors
categorical_vars <- c("nperps_cat", "attacktype1", "weaptype1", "targtype1", "suicide", "region", "country", "country_txt", "region_txt", "provstate")
impute_data[categorical_vars] <- lapply(impute_data[categorical_vars], factor)

dry_run <- mice(impute_data, maxit = 0, printFlag = FALSE)
pred_mat <- dry_run$predictorMatrix
imp_methods <- dry_run$method
imp_methods["nperps"] <- "pmm" # Use whole numbers and avoid skew
imp_methods[imp_methods != ""] <- "rf"
# If the attack was a suicide attack, we can reasonably assume the attackers is
# at least 1
imp_methods["nperps"] <- "~I(ifelse(suicide == 1 & is.na(nperps), 1, nperps))"
vars_to_exclude <- c("eventid", "country_txt", "region_txt", "provstate")
pred_mat[, vars_to_exclude] <- 0
pred_mat[vars_to_exclude, ] <- 0

# MICE with varied methods by variable
final_imputation <- mice(impute_data, 
                         m = 5, 
                         maxit = 10, 
                         method = imp_methods, 
                         predictorMatrix = pred_mat, 
                         seed = 101)

logistic_models <- with(final_imputation, 
                        glm(nperps_cat == "Lone Wolf" ~ attacktype1 + weaptype1 + targtype1 + suicide + region, 
                            family = binomial))

pooled_results <- pool(logistic_models)
```

```{r introduction_stats, include=FALSE}
attacks_over_time <- gtd_clean %>%
  #| cache: true
  filter(iyear >= 2000 & nperps >= 1) %>%
  filter(region_txt %in% c('Eastern Europe', 'Western Europe', 'North America')) %>%
  mutate(casualties = nkill + nwound) %>%
  mutate(casualties = ifelse(is.na(casualties), 0, casualties)) %>%
  arrange(iyear, imonth, iday) %>%
  group_by(iyear) %>%
  summarise(
    num_attacks = n(),
    num_indiv = sum(nperps == 1, na.rm = TRUE),
    total_casualties = sum(casualties, na.rm = TRUE),
    casualties_indiv = sum(casualties[nperps == 1], na.rm = TRUE),
    ratio = (casualties_indiv / total_casualties)
  ) %>%
  ungroup()
```

# Introduction

The term "lone wolf" was popularized in the late 1990s by white supremacists Tom Metzger and Alex Curtis as part of an encouragement to fellow racists to act alone in committing violent crimes for tactical reasons [@bakker2010lonewolves], fomenting a pivotal shift in the tactics employed of terrorism. Historically, the incidence of Lone Wolf terrorism in western countries was low in comparison to violent actors acting in groups: the lagged average of casualties perpetuated by individual terrorists was 35.4% in 2003. This proportion increased steadily for Western Countries, with the percentage rising to 85% in 2018. While analysis of recent years are still being compiled in academic datasets, initial reports show that the prior variance in this data has flattened into a continued trend: the institute for Economics and Peace reports that 93% of the fatalities in western countries in 2025 were perpetuated by lone wolves [@iep2025terrorism]. This reveals the critical need for longitudinal analysis of the evolving nature of stochastic terrorists, however the existing data is not immediately conducive to a quantitative analysis able to reliably predict tactics, outcomes and effects in its current state due to a high degree of missing fields per recorded attack.

After a careful effort of data imputation and synthesis, my analysis has indicates a statistically significant difference in attacks and methods, but no difference in the level of fatalities per attack, indicating that individuals can perpetuate violent attacks with similar outcomes to large groups, although the rate of their attacks that fail are significantly higher.

```{r intro_plot}
#| cache: true
#| label: fig1
#| fig-cap: "Attacks and casualties by year in Western Countries, 2000-2020"
plot_data <- attacks_over_time %>%
  filter(iyear != 2001) %>% # Unfortunately events in 2001 badly skew the data
  arrange(iyear) %>% 
  mutate(
    num_group = num_attacks - num_indiv,
    # Calculates a 3-year trailing average: Current year + previous year + 2 years ago
    ratio_3yr_avg = (ratio + lag(ratio, n = 1) + lag(ratio, n = 2)) / 3
  )
plot_data_long <- plot_data %>%
  dplyr::select(iyear, num_indiv, num_group) %>%
  pivot_longer(cols = c(num_indiv, num_group), 
               names_to = "attack_type", 
               values_to = "count")
scale_factor <- max(plot_data$num_attacks, na.rm = TRUE)

ggplot() +
  geom_col(data = plot_data_long, aes(x = iyear, y = count, fill = attack_type)) +
  geom_line(data = plot_data, aes(x = iyear, y = ratio_3yr_avg * scale_factor, group = 1), 
            color = "black", linewidth = 1) +
  geom_point(data = plot_data, aes(x = iyear, y = ratio_3yr_avg * scale_factor), 
             color = "black", size = 2) +
  scale_y_continuous(
    name = "Total Number of Attacks",
    sec.axis = sec_axis(~ . / scale_factor, name = "Ratio of Casualties by Individuals") 
  ) +
  
  scale_fill_manual(
    values = c("num_group" = "gray70", "num_indiv" = "steelblue"),
    labels = c("num_group" = "Group/Other", "num_indiv" = "Individuals")
  ) +
  labs(
    title = "Terrorist Attacks and Casualty Ratios in EU/NA (2000-2020)",
    subtitle = "Excluding 2001 due to extreme outlier data",
    x = "Year",
    fill = "Number of attackers"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The compound societal and economic issues inflicted upon a community by a terrorist attack are profound, leaving lasting scars that affect the stability of a region. Predicting terrorist attacks and establishing a model that connects terrorist actors to outcomes is a much-researched problem with often no clear solution [@wani2024review]. Recent breakthroughs in statistical learning methods have opened new pathways into modeling existing data sources, including more accurate data synthesis mechanism, new opportunities for the fusion and augmentation of data from external sources, and the use of contextual models to refine and establish relationships between data.

A key weakness in research into terrorism is data quality: the 'missingness' of data is an inherent issue with documenting real world events, particularly violent ones where denial and deception methods are employed to hamper law enforcement. The greatest difficulty for law enforcement is the proactive identification and capture of the "Lone Wolf" terrorist, who acts alone and does not reveal their plans before their attack. In this paper, I examine the efficacy of using Data Imputation to synthesize and classify data from the Global Terrorism Database (GTD), with a focus on identifying the number of perpetrators involved in a terrorist attack. I sought to explore the difference in lethality, tactics, and outcomes of individuals performing attacks by themselves as compared to attacks perpetuated by multiple attackers. In producing a lens through which the impact and profile of "Lone Wolves" can be approached by counterterrorism researchers, methods of identification and risk assessment can be performed. I then examined the changes in individual-sized terrorist attacks over time, looking for developing trends in attack types, success rates, and casualties, seeking to explore the rate at which these attacks are increasing in their risk.

Finally, I compare my results against the imputed dataset with the results from the limited data available in the GTD on individual attacks. While the scarcity of this data has historically presented difficulties in analysis, the value in comparing the difference in predictors from the imputed dataset reveals the confounders introduced by MICE, belying the high accuracy results of MICE's predictive model. Throughout this process I seek to ensure the ability for others to reproduce my results, as well as evaluate their validity by applying the Shapley Additive Explanation (SHAP) process to document the weights and reveal additional insight about lone wolf terrorism in comparison to other attack types.

## About the Global Terrorism Database

The systematic quantification of political violence relies upon the structural integrity of longitudinal archives, among which the Global Terrorism Database (GTD) serves as the preeminent unclassified repository for researchers, policymakers, and security analysts [@Abdalsalam2024a]. Maintained by the National Consortium for the Study of Terrorism and Responses to Terrorism (START), the GTD documents over 180,000 incidents since 1970, capturing more than 130 variables for contemporary events, ranging from tactical specifics to perpetrator identities and casualty metrics [@GTD2026]. Despite its breadth, the utility of the GTD is persistently challenged by the "missingness" problem: the presence of null or unknown entries in critical fields such as perpetrator group names, casualty counts, and specific weapon sub-types [@Abdalsalam2024a]. These informational voids are not merely random artifacts of data entry but are often the product of systematic biases in media reporting, state-sponsored censorship, or the inherent clandestine nature of non-state actors [@wani2024review].

# Prior Research

The modern frontier of data science offers a robust suite of solutions to mitigate information gaps through advanced data imputation [@Azad2025]. Traditional methods, such as listwise deletion or mean substitution, are increasingly viewed as inadequate because they frequently introduce significant bias and reduce the statistical power of subsequent analyses [@Aleryani2020]. Instead, a sophisticated integration of domain-specific Natural Language Processing (NLP) and meta-learning ensemble architectures is emerging as the gold standard for data recovery in conflict event databases. By leveraging extractive transformer models like ConfliBERT to analyze unstructured event summaries and employing Meta-Imputation Balanced (MIB) frameworks to synthesize predictions from multiple algorithmic sources, researchers have extracted contextual information by collating common attacks with synthesized data to produce models capable of highly accurately predicting group name [@Azad2025]. However, these papers and other quantitative analysis papers more interested in exploring the technical implementation and maximum efficacy of their imputation process often omit all but the largest terrorist groups, to produce groups of results that can be broadly analyzed using these contextual models [@Pan2021], focusing on groups with over 1000 documented attacks. In the case of one of Abdalsalam's papers documenting the use of the BiGRU contextual model to identify terrorist groups, Abdalsalam's dataset consisted of only terrorist groups with 500 or more documented attacks, and he also employed the SMOTE-T data resampling technique to counter minority bias from underrepresented groups within this dataset of only 36 terrorist groups, demonstrating the confounding effect of missingness of the GTD's dataset on some analysis and imputation techniques [\@ @Abdalsalam2024a's].

Wani explores the mechanisms of missing data completely at random (MCAR), when the observed data is unevenly missing at random (MAR) and not missing at random (NMAR), in which the missingness is a result of a confounding effect [@wani2024review]. In their review of potential imputation frameworks, Wani evaluates recovery strategies such as Multiple Imputation by Chained Equations (MICE), Random Forests, and Generative Adversarial Networks. Wani advocates for task-specific evaluation metrics rather than relying solely on standard error calculations to address imbalanced NMAR datasets such as the GTD which exhibit high dimensionality. In datasets such as these, fields are left blank or intentionally flagged as unknown if there is not clear consensus, even if some potential information points toward a value. Indeed, Abdasalam notes in their 2024 article exploring the missingness of the group name field that a value being left unknown is sometimes a conscious choice, stating "This gap in research may stem from the complex and dynamic nature of terrorist groups, which frequently evolve, change names, merge, or disband, thereby complicating the creation of accurate and current classification and prediction models." [@Abdalsalam2024a].

Reductions in the analyzed dataset have also prevailed to produce statistically significant predictions at the cost of external validity. Abdalsalam has had success with highly predictive contextual prediction models such as DistilBERT [@Abdalsalam2021a] or extractive text classifiers such as word2vec [@Abdalsalam2024], resulting in models capable of identifying attack type or location with a high degree of accuracy, although the research did exclude about 73% of the dataset to missing or insufficiently long descriptions. This exposes the limitations of predicting a 'freeform' field such as group name using a statistical model without a limited Retrieval Augmented Generation (RAG) ability to narrow the search space of the contextual model. In the tangential field of Biomedical data, Azad has shown the promise of ensemble-based methods that combine contextual models with ensemble models to produced balanced approaches that adjust weights of predictors across multiple missing fields [@Azad2025]. This approach may allow analysts to bridge the gap between records where generating a RAG of applicable context (such as English language news articles) is not available in a timely

Accounting for the level of missingness using a mask indicator is also useful to identify patterns in NMAR datasets. Using this method, Perez-Lebel benchmarked imputation approaches on healthcare databases and found that tree-based models utilizing Missing Incorporated in Attributes (MIA) provide the best balance of accuracy and speed [@PerezLebel2022]. When data is weakly correlated, a Ranked k-Nearest Neighbors (RkNN) can perform well. @Khan2024 introduces the RkNN algorithm to refine traditional kNN imputation. By utilizing Ranked Set Sampling, this algorithm better handles data variability and reduces error when variables are weakly correlated or lack sufficient nearby neighbors.

Extensive research has been performed leveraging ensemble frameworks to reduce variance and improve prediction accuracy. @Aleryani2020 proposed Multiple Imputation Ensembles (MIE), a system where multiple imputation techniques and ensemble methods collaborate to produce a superior final prediction when data is Missing Completely At Random. @Azad2025 detailed the Meta-Imputation Balanced (MIB) framework, which trains a supervised meta-model on synthetically masked data to intelligently weigh the outputs of multiple base imputers. These methods are automated, powerful tools, which trade computational performance for accuracy. A key advantage is the ability for the models to seek and identify statistically significant fields on their own, although this can also be a hazard when colinear variables or keys are included for contention, reducing the external validity of the model to impute new data.

The authors above have all successfully used ensemble-based models of various flavors to produce highly accurate predictive models. However, maintaining external validity and reproducability of the model is also critical. Štrumbelj and Kononenko discuss the process of evaluating the impact of ensemble models in the final product by adapting the Shapley value. The authors quantify the "fair" contribution of each feature value to a specific classification outcome, effectively revealing how much each variable influences a model's decision. To resolve the exponential time complexity typically associated with these calculations, the paper proposes a sampling-based approximation method that makes the technique computationally feasible for complex datasets [@JMLR:v11:strumbelj10a]. New compound fields can be generated against the GTD by analyzing text fields alongside restricting datasets to natural groups. Bridgelall describes an empirical text mining algorithm to produce a new "motive" field that allows analysts to categorize attacks based on underlying sentiments such as anti-government "rule," race, or religion. The study employs advanced techniques like k-means co-clustering and nonlinear projections to visualize how these motives statistically intersect with specific tactics and targets [@bridgelall2022unsupervised]. This can allow the conversion of freeform and numerical fields into factorized categories, allowing for additional analysis pathways.

Another option for field synthesis is the use of additional data sources connected without the direct tie between two fields. @Cai2025 developed a sophisticated AI-driven tool called FAME (Fingerprint to Article Matching for Events) to systematically track how global media outlets report on natural disasters and terrorist attacks. By utilizing a "fingerprint" constructed from a specific combination of event class, location, and date, the system filters millions of multilingual articles to identify those covering specific crises without the need for manual training data. This methodology reveals that news coverage is heavily biased toward wealthier nations and those with significant trade ties, often requiring much higher death tolls in poorer regions to receive equivalent attention. Ultimately, this work provides an open-source framework for computational social science to measure and understand the systemic factors that shape the global news agenda.

## Hypothesis

The result of this robust analysis into the methods of accurately synthesizing data opens an opportunity to impute the number of perpetrators for a large slice of the GTD, omitting only the largest number of attack records to curb outliers. In predicting the number of perpetrators for every remaining record, an analysis of attacks predicted to be perpetuated by individuals will both expose recent trends on attack patterns and tactics, but also allow for a broader analysis of terrorism trends in regions such as North America and Europe which currently have a comparatively small number of records to quantitatively predict from. It is my prediction that the analysis of imputed data in comparison to the original dataset will reveal the confounding effect of the NMAR number of perpetrators to indicate which attack and weapon types have the highest degrees of casualty-inducing outcomes.

# Process and Results

High-quality, varied data is the often-sought, rarely-acquired state for quantitative analysis. To limit the scope of my longitudinal analysis to roughly the present day conditions, I reduced the records of the GTD dataset to only records from 2000 to 2020, and I also excluded records with 50 or more perpetrators, as these records heavily skew the data. After the dataset had been scoped, I examined the remaining 138,000 records. Of these, 22,065 records (15.9%) had a defined nperps value, with 8,520 (5.9%) records having a single attacker. These rows comprise my test and training dataset.

The high degree of missing fields reflects a strategy by the GTD researchers at START to only include data they can establish with absolute certainty [@GTD2007introducing]. As a result, while information on attackers and outcomes might be fairly robust, if the accounts of an attack only refer to "a handful of attackers" or "an unknown assailant," it is likely that the number of perpetrators would be left blank. I hypothesize that for the purposes of statistical analysis of outcomes, defining the exact number of actors is likely required; rather, establishing broader categories of Individual, Small Group and Large Group would be sufficient to convert the number of perpetrators into a categorical factor which could be used for imputation and prediction. In this way, the accuracy and specificity of the imputation model is greatly increased, without compromising the field for my purposes. By using Multivariate imputation by chained equations (MICE), I applied this method on the GTD dataset, with a few caveats beforehand: I assumed that attacks marked as suicide attacks in the GTD involved at least one perpetrator: The number of perpetrators must at least be nonzero in this case. I also restricted the variables used in my imputation model to primary attack type (attacktype1), primary weapon type (weaptype1), primary target (targtype1), and the broader region and country variables. I did not restrict my model to a specific time or country bounds, in the hopes that the resulting model would retain its validity independent of longitudinal or location confounding effects.

Examining the GTD data and existing research, I identified that the GTD data spanning from 2000-2020 has a number of critical fields such as Number of perpetrators, Group name and casualty information that have a high level of missingness. These fields are my primary candidates for imputation, and group name in particular has value in identifying not only groups involved, but also in establishing which attacks currently defined as "unknown" could instead be attributed to independent "Lone Wolf" attackers, the propensity of which is increasing over time and represents a significant threat for counterterrorism law enforcement [@bakker2010lonewolves].

```{r gtd_plot}
#| cache: true
#| label: fig2
#| fig-cap: "Select fields ordered by their percentage of missingness in the GTD"
# r-graph-gallery is a lifesaver here
plot_data <- gtd_missing_data_summary %>%
  mutate(
    clean_name = case_match(
      Column,
      "gname" ~ "Group Name",
      "nperps" ~ "Number of Perpetrators",
      "nkill" ~ "Number Killed",
      "nwound" ~ "Number Wounded",
      "motive" ~ "Motive",
      .default = Column # Keeps the original name if it doesn't match above
    ),
    # If the bar is > 20%, put the name inside. If not, leave it blank ("").
    inside_label = ifelse(Missing_Pct > 20, clean_name, ""),
    outside_label = ifelse(Missing_Pct > 20, 
                           sprintf("%.1f%%", Missing_Pct), 
                           paste0(clean_name, " (", sprintf("%.1f%%", Missing_Pct), ")"))
  )

ggplot(plot_data, aes(x = Missing_Pct, y = reorder(clean_name, Missing_Pct))) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(x = 1, label = inside_label), 
            hjust = 0, color = "white", size = 3.5, fontface = "bold") +
  geom_text(aes(label = outside_label), 
            hjust = -0.1, color = "black", size = 3.5) +
  scale_x_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.4))) +
  labs(
    title = "GTD: Percentage of Missing Data by Column",
    subtitle = "Includes NA values and 'Unknown' strings",
    x = "Missing Percentage (%)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

In an opportunity to provide additional context to the GTD, other adjacent datasets on terrorism can be compared. The Armed Conflict Location and Event Data (ACLED) database provides valuable context to the GTD when collated by location and time. The ACLED includes details on news sources describing the attribution [@ACLEDcodebook2025] information as well as detailed actor information. While the ACLED contains nearly significantly more entries for the same time period, it is important to note the difference in membership criterion. The GTD contains information on confirmed non-state terrorist attacks where money was not the primary goal of the attack [@GTD2007introducing]; the ACLED has a much broader selection criterion including state actors and independent militias, as well as protest and riots that would be excluded from the GTD. A joined database of shared records brings the intersecting rows down to 20,782 records, removing 118,850 records from the potential comparison pool. As a result, my imputation workflow will be run on the standalone GTD dataset as well as a joined dataset, giving opportunities for accuracy comparisons between the two, holding the workflow process consistent.

```{r acled_plot}
#| label: fig3
#| fig-cap: "Select fields ordered by their percentage of missingness in the ACLED"

ggplot(acled_missingness_summary, aes(x = Column, y = Missing_Pct)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", Missing_Pct)), 
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.1))) +
  scale_x_discrete(labels = aclet_target_col_labels) +
  labs(
    title = "ACLED: Percentage of Missing Data by Column",
    subtitle = "Includes NA values and 'Unknown' strings",
    x = "Column Name",
    y = "Missing Percentage (%)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
#rm(acled_clean, acled_missingness_summary)
```

My research indicated that Multivariate imputation by chained equations (MICE) was the right combination of performance and accuracy to reliably perform data imputation on the number of perpetrators `nperps` field, but the efficacy of the model greatly improves if the range of predicted options is more narrow than a numerical value. As a result, I categorized the number of perpetrators into 3 groups: individuals, small groups (2-5 actors), and large groups (6+). I then used the resulting model from the MICE process to impute the categorical factor of perpetrators for all records in my dataset.

## Accuracy of MICE imputation against the GTD

After training, I generated a confusion matrix against the unimputed data for comparison purposes. The results of the MICE model reflect a relatively high degree of accuracy in predicting if an attack was perpetrated by an individual (97.7%), with decreasing performance for the small and large group categories. I then applied logistic regression against my imputed results to examine the differences in outcomes and tactics, if any, between attacks by individuals and groups. I determined that while individuals and groups have similar outcomes when the same attack and weapon types are considered, individual terrorist attackers are far more likely to employ vehicles and to commit suicide attacks.

```{r mice_accuracy, include=FALSE}
#| cache: true
# Map over final methodologies
pred_mat <- final_imputation$predictorMatrix
imp_methods <- final_imputation$method

# Confusion matrix of MICE process against nperps_cat
truth_data <- impute_data %>% filter(!is.na(nperps_cat))
set.seed(101)
masked_data <- truth_data
mask_indices <- sample(1:nrow(masked_data), size = floor(0.3 * nrow(masked_data)))
true_categories <- masked_data$nperps_cat[mask_indices]
masked_data$nperps_cat[mask_indices] <- NA

eval_imputation <- mice(masked_data, 
                        m = 1, # 1 dataset is sufficient for the evaluation run
                        maxit = 5, 
                        method = imp_methods, 
                        predictorMatrix = pred_mat, 
                        seed = 123)
rm(dry_run)
```

```{r conf_table}
#| cache: true
#| label: fig4
#| fig-cap: "Performance Metrics of MICE Imputation by Category"
completed_eval_data <- complete(eval_imputation, 1)
imputed_categories <- completed_eval_data$nperps_cat[mask_indices]

# Generate the Confusion Matrix
# It is critical to ensure both vectors have the exact same factor levels
imputed_categories <- factor(imputed_categories, levels = c("Lone Wolf", "Small Cell", "Large Group"))
true_categories <- factor(true_categories, levels = c("Lone Wolf", "Small Cell", "Large Group"))
conf_matrix <- confusionMatrix(data = imputed_categories, reference = true_categories)

# Extract Metrics and Format Summary Table
# Extracting Sensitivity and Specificity, and calculating the percentage difference 
# from the previous row based on Sensitivity.
metrics_table <- as.data.frame(conf_matrix$byClass[, c("Sensitivity", "Specificity")])
metrics_table$Class <- gsub("Class: ", "", rownames(metrics_table))
metrics_table <- metrics_table %>%
  mutate(
    `Pct Diff in Sensitivity from Prev Row` = sprintf("%.2f%%", (Sensitivity - lag(Sensitivity)) / lag(Sensitivity) * 100)
  ) %>%
  dplyr::select(Class, Sensitivity, Specificity, `Pct Diff in Sensitivity from Prev Row`)

metrics_table <- metrics_table %>%
  mutate(
    `Pct Diff in Sensitivity from Prev Row` = replace_na(`Pct Diff in Sensitivity from Prev Row`, "-")
  )
    
# Generate the table
metrics_table %>%
  kbl(
    caption = "Performance Metrics of MICE Imputation by Category",
    booktabs = TRUE, # Essential for the clean, academic look (LaTeX/PDF)
    digits = 3,
    align = c("l", "c", "c", "c", "C")
  ) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  add_header_above(c(" " = 1, "Performance Metrics" = 4)) %>%
  footnote(general = "Pct Diff compares the sensitivity of the current row to the row above it.")
```

```{r conf_plot}
#| cache: true
#| label: fig5
#| fig-cap: "Confusion Matrix for Imputed Categories"
cm_df <- as.data.frame(conf_matrix$table)
ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "black", alpha = 0.8) +
  geom_text(aes(label = Freq), size = 6, color = ifelse(cm_df$Freq > mean(cm_df$Freq), "white", "black")) +
  scale_fill_gradient(low = "white", high = "#2C3E50") +
  theme_minimal() +
  labs(
    title = "Confusion Matrix for Imputed Categories",
    x = "True Category (Reference)",
    y = "Imputed Category (Prediction)",
    fill = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold")
  )
rm(eval_imputation, masked_data, truth_data, completed_eval_data)
```

```{r mice_results_plot}
#| cache: true
#| label: fig6
#| fig-cap: "Predictors of Independent Actor (Lone Wolf) Terrorism"
# Tidy the data and calculate Odds Ratios at 95% confidence
plot_data <- tidy(pooled_results, conf.int = TRUE, exponentiate = TRUE)
plot_data <- plot_data %>%
  # Remove the Intercept, as it only represents the baseline odds and skews the plot
  filter(term != "(Intercept)") %>%
  mutate(
    category = case_when(
      grepl("attacktype", term) ~ "Attack Type",
      grepl("suicide", term) ~ "Attack Type",
      grepl("weaptype", term) ~ "Weapon Type",
      grepl("targtype", term) ~ "Target Type",
      grepl("region", term) ~ "Region",
      TRUE ~ "Other"
    ),
    clean_term = case_when(
# Attack Types 
      term == "attacktype11" ~ "Assassination",
      term == "attacktype12" ~ "Armed Assault",
      term == "attacktype13" ~ "Bombing/Explosion",
      term == "attacktype14" ~ "Hijacking",
      term == "attacktype15" ~ "Hostage Taking (Barricade)",
      term == "attacktype16" ~ "Hostage Taking (Kidnapping)",
      term == "attacktype17" ~ "Facility/Infrastructure Attack",
      term == "attacktype18" ~ "Unarmed Assault",
      term == "attacktype19" ~ "Unknown",
      term == "suicide1" ~ "Suicide Attack",
      
      # Weapon Types
      term == "weaptype11" ~ "Biological",
      term == "weaptype12" ~ "Chemical",
      term == "weaptype13" ~ "Radiological",
      term == "weaptype14" ~ "Nuclear",
      term == "weaptype15" ~ "Firearms",
      term == "weaptype16" ~ "Explosives",
      term == "weaptype17" ~ "Fake Weapons",
      term == "weaptype18" ~ "Incendiary",
      term == "weaptype19" ~ "Melee",
      term == "weaptype110" ~ "Vehicle",
      term == "weaptype111" ~ "Sabotage Equipment",
      term == "weaptype112" ~ "Other",
      term == "weaptype113" ~ "Unknown",
      
      # Target Types
      term == "targtype11" ~ "Business",
      term == "targtype12" ~ "Government (General)",
      term == "targtype13" ~ "Police",
      term == "targtype14" ~ "Military",
      term == "targtype15" ~ "Abortion Related",
      term == "targtype16" ~ "Airports & Aircraft",
      term == "targtype17" ~ "Government (Diplomatic)",
      term == "targtype18" ~ "Educational Institution",
      term == "targtype19" ~ "Food or Water Supply",
      term == "targtype110" ~ "Journalists & Media",
      term == "targtype111" ~ "Maritime",
      term == "targtype112" ~ "NGO",
      term == "targtype113" ~ "Other",
      term == "targtype114" ~ "Private Citizens & Property",
      term == "targtype115" ~ "Religious Figures/Institutions",
      term == "targtype116" ~ "Telecommunication",
      term == "targtype117" ~ "Terrorists/Non-State Militia",
      term == "targtype118" ~ "Tourists",
      term == "targtype119" ~ "Transportation",
      term == "targtype120" ~ "Unknown",
      term == "targtype121" ~ "Utilities",
      term == "targtype122" ~ "Violent Political Parties",

      # Regions 
      term == "region1" ~ "North America",
      term == "region2" ~ "Central America & Caribbean",
      term == "region3" ~ "South America",
      term == "region4" ~ "East Asia",
      term == "region5" ~ "Southeast Asia",
      term == "region6" ~ "South Asia",
      term == "region7" ~ "Central Asia",
      term == "region8" ~ "Western Europe",
      term == "region9" ~ "Eastern Europe",
      term == "region10" ~ "Middle East & North Africa",
      term == "region11" ~ "Sub-Saharan Africa",
      term == "region12" ~ "Australasia & Oceania",
      
      TRUE ~ term 
    )
  ) %>%
  # Filter out the Radiological outlier causing the X-axis blowout
  # "Radiological weapons were excluded from the logistic regression visualization due to quasi-complete separation caused by insufficient event frequency
  filter(clean_term != "Radiological")

forest_plot_quad <- ggplot(plot_data, aes(x = estimate, y = reorder(clean_term, estimate), color = category)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "darkgrey", size = 1) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "#2C3E50", size = 0.8) +
  geom_point(size = 3, color = "#E74C3C") +
  scale_x_log10() + # Use a logarithmic scale for the X-axis
  facet_wrap(~ category, scales = "free_y", ncol = 2) + # 4 times the chart, bruv
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Predictors of Independent Actor (Lone Wolf) Terrorism",
    subtitle = "Odds Ratios with 95% Confidence Intervals (Pooled from MICE)",
    x = "Odds Ratio (Log Scale)",
    y = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "none"
  )

forest_plot_quad
#ggsave("forest_plot_lone_wolf.png", plot = forest_plot_quad, width = 12, height = 10, dpi = 300)
```

```{r complete_full_data, include=FALSE}
#| cache: true
# Extract all 5 models, then average and consolidate, to attempt to reduce variance
imputed_long <- complete(final_imputation, action = "long")
imputed_averaged <- imputed_long %>%
  group_by(eventid) %>%
  summarise(
    # Calculate the mean and round to the nearest whole number
    nperps_avg = round(mean(nperps)),
    nkill_avg = round(mean(nkill)),
    nwound_avg = round(mean(nwound))
  ) %>%
  ungroup() %>%
  # Rebuild nperps_cat
  mutate(
    nperps_cat = case_when(
      nperps_avg == 1 ~ "Lone Wolf",
      nperps_avg > 1 & nperps_avg <= 5 ~ "Small Cell",
      nperps_avg > 5 ~ "Large Group",
      TRUE ~ NA_character_
    ),
    nperps_cat = as.factor(nperps_cat)
  )

gtd_clean_final <- gtd_clean %>%
  dplyr::select(-c(nperps, nperps_cat, nkill, nwound)) %>%
  left_join(imputed_averaged, by = "eventid") %>%
  rename(
    nperps = nperps_avg,
    nkill = nkill_avg,
    nwound = nwound_avg
  )

# Verify the final structure
str(gtd_clean_final %>% dplyr::select(eventid, nperps, nperps_cat, nkill, nwound))
rm(imputed_long, imputed_averaged, impute_data)
```

```{r nperps_on_nkill, include=FALSE}
#| cache: true
# Evaluate the effects of perpetrator group size on the number of individuals killed in the attack
kruskal.test(nkill ~ nperps_cat, data = gtd_clean_final)
# Result is near-zero, so what is the relationship between category and nkill?
pairwise.wilcox.test(gtd_clean_final$nkill, gtd_clean_final$nperps_cat,
                     p.adjust.method = "holm")
```

A Kruskal-Wallis rank sum test was conducted to determine if there were differences in the number of casualties (nkill) based on the size of the perpetrator group. The results indicated a highly significant difference in the distribution of casualties between Lone Wolf, Small Cell, and Large Group attacks ($H(2) = 1136.6, p < 0.001$). To determine where the significant differences in casualties lay among the perpetrator categories, a post-hoc pairwise Wilcoxon rank sum test was conducted using the Holm method to control for multiple comparisons. The analysis revealed that Lone Wolf attacks resulted in significantly different casualty distributions when compared to both Small Cells ($p < 0.001$) and Large Groups ($p < 0.001$). Notably, there was no statistically significant difference in fatalities between Small Cells and Large Groups ($p = 0.53$). This suggests a critical threshold effect in terrorist operations: while independent actors exhibit a distinct lethality profile, the introduction of even a small network (2-5 perpetrators) shifts the casualty distribution to match that of large-scale group attacks. Consequently, casualty metrics serve as a strong discriminator for imputing independent actors, but offer diminishing returns when attempting to differentiate the size of multi-perpetrator networks.

When examining the effect that tactics and group composition has on casualities in terror attacks, the extreme overdispersion of casualties means that a standard regression underestimates the standard errors, leading to false significance. This is a particular risk when considering the imputed dataset. As a result, I employed a Zero-inflated negative binomial (ZINB) model to examine the relationship between nkill and the other predictors. The ZINB model seperates the structurally missing (data not determined by GTD) data from the attack data where the attack was a failure. However, upon production of the model, its predictors bias matched the Negative Binomial result.

```{r zinb_model, include=FALSE}
#| cache: true
gtd_clean_final$nperps_cat <- relevel(gtd_clean_final$nperps_cat, ref = "Lone Wolf")

zinb_models <- with(gtd_clean_final, 
                    zeroinfl(nkill ~ nperps_cat + region | attacktype1 + weaptype1, 
                             dist = "negbin"))
```

```{r rootogram_plot_3, include=FALSE}
#| cache: true
plot_nb <- glm.nb(nkill ~ nperps_cat + region + attacktype1 + weaptype1, 
                  data = gtd_clean_final)

# Unfortunately these zinb calculations don't render. However the result matches
# the nb plot identically.
#plot_zinb <- zeroinfl(nkill ~ nperps_cat + region | attacktype1 + weaptype1, 
#                      data = gtd_clean_final, dist = "negbin")
# ZINB calculations
#expected_probs_zinb <- predict(plot_zinb, type = "prob")
#expected_freq_zinb <- colSums(expected_probs_zinb[, 1:(plot_limit + 1)])
```

```{r rootogram_plot_4, include=FALSE}
# Process the Data (Filtering to 25 to keep the plot readable)
plot_limit <- 25 

# Standard NB calculations
mu_nb <- predict(plot_nb, type = "response")
theta_nb <- plot_nb$theta
expected_freq_nb <- sapply(0:plot_limit, function(i) sum(dnbinom(i, mu = mu_nb, size = theta_nb)))
observed_freq <- table(factor(gtd_clean_final$nkill, levels = 0:plot_limit))

df_nb <- data.frame(
  Count = 0:plot_limit,
  Observed = as.numeric(observed_freq),
  Expected = as.numeric(expected_freq_nb)
) %>%
  mutate(Sqrt_Obs = sqrt(Observed), Sqrt_Exp = sqrt(Expected), Bottom = Sqrt_Exp - Sqrt_Obs)

#df_zinb <- data.frame(
#  Count = 0:plot_limit,
#  Observed = as.numeric(observed_freq),
#  Expected = as.numeric(expected_freq_zinb)
#) %>%
#  mutate(Sqrt_Obs = sqrt(Observed), Sqrt_Exp = sqrt(Expected), Bottom = Sqrt_Exp - Sqrt_Obs)
```

```{r rootogram_plot_5, include=FALSE}
#| cache: true
# Helper function
build_rootogram <- function(data, title) {
  ggplot(data, aes(x = Count)) +
    geom_segment(aes(x = Count, xend = Count, y = Sqrt_Exp, yend = Bottom), 
                 linewidth = 4, color = "steelblue", alpha = 0.8) +
    geom_line(aes(y = Sqrt_Exp), color = "red", linewidth = 1) +
    geom_point(aes(y = Sqrt_Exp), color = "red", size = 2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
    labs(
      title = title,
      x = "Number of Fatalities (nkill)",
      y = expression(sqrt("Frequency"))
    ) +
    theme_minimal() +
    theme(plot.title = element_text(face = "bold", size = 12))
}
```

```{r rootogram_plot_6}
#| cache: true
#| label: fig7
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 3
#| fig-cap: "Rootogram comparison demonstrating model fit. The Zero-Inflated model and the Standard Negative Binomial exhibit similar performance"
#| fig-align: "center"
# Create both plots
plot1 <- build_rootogram(df_nb, "Standard Negative Binomial")
#plot2 <- build_rootogram(df_zinb, "Zero-Inflated Negative Binomial")
plot1 # | plot2 (ZINB identical to NB, can omit)
```

```{r zinb_model_2, include=FALSE}
#| cache: true
# Blessed be stack overflow, praise always to those who have figured this out before me
# Define the missing tidy method for zeroinfl models
tidy.zeroinfl <- function(x, conf.int = FALSE, conf.level = 0.95, ...) {
  s <- summary(x)
  
  # Extract Count Model coefficients and PREFIX the term names
  count_df <- as.data.frame(s$coefficients$count)
  count_df$term <- paste0("count_", rownames(count_df)) 
  
  # Extract Zero Model coefficients and PREFIX the term names
  zero_df <- as.data.frame(s$coefficients$zero)
  zero_df$term <- paste0("zero_", rownames(zero_df))
  
  # Combine
  df <- rbind(count_df, zero_df)
  rownames(df) <- NULL
  
  # Standardize for mice
  colnames(df)[1:4] <- c("estimate", "std.error", "statistic", "p.value")
  
  return(df)
}

# Extract the results with confidence intervals
#pooled_zinb_results <- pool(zinb_models)
#zinb_analysis_raw <- tidy(pooled_zinb_results, conf.int = TRUE)

# View the raw summary
#summary(pooled_zinb_results)
# Extract the results
#zinb_analysis <- tidy(pooled_zinb_results, conf.int = TRUE)
# Clean up the names and recreate the 'component' column
#zinb_analysis <- zinb_analysis_raw %>%
#  mutate(
    # Create the component column based on the prefix
#    component = ifelse(grepl("^count_", term), "count", "zero"),
    # Strip the prefix out of the term name
#    term = gsub("^count_|^zero_", "", term)
#  )

# Separate the models for easier exponentiation and reading
#count_model_results <- zinb_analysis %>% filter(component == "count")
#zero_model_results <- zinb_analysis %>% filter(component == "zero")

# Exponentiate the count model to get Incidence Rate Ratios (IRRs)
#count_model_irrs <- count_model_results %>%
#  mutate(estimate_irr = exp(estimate),
#         conf.low_irr = exp(conf.low),
#         conf.high_irr = exp(conf.high))

# Exponentiate the zero model to get Odds Ratios (ORs)
#zero_model_ors <- zero_model_results %>%
#  mutate(estimate_or = exp(estimate),
#         conf.low_or = exp(conf.low),
#         conf.high_or = exp(conf.high))
```

\clearpage

# Annotated Bibliography

*Note: I have done extensive searching for related works, background methods, and a variety of techniques. I think it's unlikely that I will use all of these references in the final paper, but I had to undertake significant research to understand the most recent breakthroughs in data imputation and studies against the GTD.*

## **Background References**

This research brief from the National Consortium for the Study of Terrorism and Responses to Terrorism (START), titled "**Anti-State Terrorist Plots and Attacks in the United States**" examines a significant escalation in anti-state plots and attacks within the United States between 1992 and 2022. By analyzing data from the PIRUS project, the authors reveal that anti-government extremists have driven recent violence to historical highs, increasingly through lone-actor operations rather than organized groups. The text highlights a unique profile for these offenders, noting they are typically older, more socially established, and possess military backgrounds at rates far exceeding the general population [@jensen2024antistate].

This research paper by the International Centre for Counter-Terrorism investigates the lone wolf phenomenon, defined as individuals who commit politically or religiously motivated violence without formal ties to a parent organization. The article, titled "**Lone Wolves: How to prevent this phenomenon?"** describes a crucial confounder in the GTD and many papers analyzing its data, which is that individual attacks from lone wolves aren't able to be easily predicted using the terrorist-group focused fields of the dataset. The authors explore how these actors pose a unique detection challenge for intelligence services because they lack a communicative network, even though they often share common psychological traits and ideological inspirations found on the internet. By analyzing the tactical shift toward leaderless resistance across various extremist backgrounds, the text seeks to move beyond traditional profiling by focusing on the operational behaviors and radicalization triggers of these solitary attackers [@bakker2010lonewolves].

## **Contextual References**

This paper from 2024 titled "**A Review of Challenges and Solutions for Using Machine Learning Approaches for Missing Data**" explores the common problems encountered by researchers handling holes in datasets, and what impacts mitigation strategies have on outcomes. The author categorizes the mechanisms of missingness---MCAR, MAR, and NMAR---to provide a conceptual framework for selecting appropriate recovery strategies like MICE, Random Forest, and Generative Adversarial Networks. Beyond simple techniques, the text explores critical hurdles such as handling heterogeneous data, managing high dimensionality, and addressing imbalanced datasets where minority classes are at risk of bias. To ensure data integrity, the paper advocates for task-specific evaluation metrics and robustness testing rather than relying solely on standard error calculations [@wani2024review].

Examining "**A novel ranked k-nearest neighbors algorithm for missing data imputation",** This research paper introduces the Ranked k-Nearest Neighbors (RkNN) algorithm, a refinement of traditional machine learning techniques designed to estimate missing information in complex datasets. While the standard kNN approach relies on simple proximity to fill gaps, it can fail when variables are weakly correlated or when only a few neighbors are in range. The authors are better able to handle data variability and reduce error via their implementation of Ranked Set Sampling. Extensive testing on both simulated environments and real-world economic data demonstrates that this new algorithm provides superior imputation accuracy across various types of data loss [@Khan2024].

"**Benchmarking missing-values approaches for predictive models on health databases**" presents a comprehensive systematic benchmark evaluating how different machine learning strategies handle missing values in large-scale healthcare databases. By analyzing thirteen prediction tasks across four major health datasets, the authors conclude that tree-based models utilizing Missing Incorporated in Attributes (MIA) offer the best balance of accuracy and speed. While complex multiple imputation and bagging can slightly improve performance, they are often computationally prohibitive compared to the efficiency of native support for missing data. The study highlights that adding a missingness indicator or "mask" significantly boosts accuracy, suggesting that the absence of data in medical records often contains informative patterns that are crucial for predictive modeling [@PerezLebel2022].

In the "**Practical guide to SHAP analysis: Explaining supervised machine learning model predictions in drug development**", The authors describe the essential requirement of SHapley Additive exPlana-tions (SHAP), a powerful tool designed to demystify the "black-box" nature of machine learning models. SHAP allows authors to describe their process and improve reproducibility, which will be essential for my research replicating prior imputation methods used against the GTD. The process relies on dividing and representing the impact of each workflow step's incremental impact on the final model's predictive power or methodology [@PonceBobadilla2024].

On the subject of data imputation, this article from an adjacent dataset titled "**Final Report: Empirical Assessment of Domestic Radicalization (EADR)**" provides value on handling data imputation and prediction. The report, funded by the National Institute of Justice, provides an empirical assessment of domestic radicalization in the United States by utilizing the PIRUS database to analyze the backgrounds and behaviors of extremists. The study employs a dual-methodological approach, combining multivariate regression models with fuzzy-set Qualitative Comparative Analysis (fs/QCA) to identify the specific causal pathways that lead individuals toward violent versus non-violent extremism. A significant portion of the text is dedicated to rigorous data calibration and missing data strategies, ensuring that the findings regarding factors like criminal history, mental health, and social cliques are statistically robust [@jensen2016final].

The paper "**Multi-behavioral Sequential Prediction with Recurrent Log-bilinear Model**" attempts to address the complexity of how users interact with items over time. Traditional models often struggle to differentiate between various actions such as downloading versus uninstalling an app, and fail to capture the non-monotonic influence of past behaviors on future choices. To solve this, the authors introduce the Recurrent Log-Bilinear (RLBL) model and its time-aware variant (TA-RLBL), which use position-specific and behavior-specific matrices to better represent short-term and long-term context. By incorporating temporal differences and specialized transition matrices, these models significantly outperform standard methods in accurately forecasting a user's next specific action across diverse real-world datasets [@Liu2016].

Another imputation pathway would be the use of Multiple Imputation Ensembles (MIE). Aleryani et. all's paper "**Multiple Imputation Ensembles (MIE) for dealing with missing data**" describes a robust framework designed to improve the accuracy of machine learning classification when datasets contain missing data. In merging multiple imputation techniques with ensemble methods, the authors create a system where multiple models collaborate to produce a superior final prediction. The study rigorously tests this approach against traditional strategies, such as simple imputation and internal algorithm mechanisms, using twenty diverse benchmark datasets with varying levels of data Missing Completely At Random (MCAR). To assess quality, the researchers utilize dissimilarity measures to determine how closely the imputed values resemble the original information [@Aleryani2020].

## **Prior Work References**

The Meta-Imputation Balanced (MIB) framework shows a lot of promise for missing data generation. This paper, titled "**Meta-Imputation Balanced (MIB): An Ensemble Approach for Handling Missing Data in Biomedical Machine Learning**" describes how MIB functions as an ensemble approach that intelligently combines the outputs of multiple base imputers using a supervised meta-model. By training on data where values have been intentionally hidden, the system learns to weigh different imputation strategies to produce the most accurate and stable estimates for the missing entries. Experimental results across various biomedical datasets demonstrate that this context-aware strategy consistently achieves high accuracy and maintains strong downstream predictive performance, proving more reliable than individual traditional or modern methods [@Azad2025].

Another option for Imputation is DeepIFSAC or GAIN, which this article contrasts alongside other techniques. Titled "**DeepIFSAC: Deep Imputation of Missing Values Using Feature and Sample Attention within Contrastive Framework**" the article describes a sophisticated deep learning framework designed for missing value imputation in tabular datasets where data is often incomplete. The model utilizes a novel joint attention mechanism that simultaneously analyzes relationships between features (columns) and between samples (rows) to accurately estimate missing information. To enhance robustness, the authors incorporate CutMix data augmentation and contrastive learning, which helps the system distinguish similar data patterns and better handle non-random missing data. Extensive benchmarking across diverse datasets demonstrates that this method consistently outperforms state-of-the-art statistical and machine learning models, particularly when missingness is high [@Kowsar2025].

The 2021 paper is part of a series by Mohammed Abdalasam. "**A Study of the Effects of Textual Features on Prediction of Terrorism Attacks in GTD Dataset"** examines the textual fields within the GTD to make predictions on other fields, specifically attack types. The authors propose a machine learning framework that combines standard categorical and numerical"key features" with descriptive text summaries of incidents. To process this unstructured text, they utilized various representation techniques, including TF-IDF, Bag of Words, and Word Embedding, and evaluated the performance of nine different machine learning classifiers. Incorporating these textual features significantly enhances accuracy, with the ensemble Bagging classifier and Word2vec embeddings achieving the most effective results in identifying attack patterns [@Abdalsalam2021a].

Abdalasam's research using the GTD continues in "**Terrorism Attack Classification Using Machine Learning**" to refine the approach of harvesting the textual fields of the dataset to make predictions. The authors argue that traditional analysis often neglects the rich information found in textual summaries within the Global Terrorism Database (GTD), which can significantly enhance a model's predictive accuracy. To bridge this gap, they utilize DistilBERT, a transformer-based model, to extract deep contextual and semantic features from raw text descriptions of incidents. These extracted textual features are then integrated with standard numerical and categorical data to train various machine learning classifiers. The study concludes that this hybrid feature approach dramatically outperforms models relying on single data types, achieving a peak classification accuracy of 98.7% using the extreme gradient boosting (XGBoost) algorithm [@Abdalsalam2024].

Abdalasam then used BiGRU-SA, a sophisticated artificial intelligence framework designed to accurately identify the specific terrorist organizations responsible for violent attacks. In their article "**Terrorism group prediction using feature combination and BiGRU with self-attention mechanism**", the authors combine natural language processing via DistilBERT to analyze incident summaries in the GTD with advanced data balancing techniques like SMOTE-T to ensure fair representation of smaller groups. The study demonstrates that their model, which integrates bidirectional gated recurrent units and self-attention mechanisms, achieves an exceptional 98.68% accuracy rate, significantly outperforming traditional machine learning methods [@Abdalsalam2024a].

The research paper "**Predicting The Unpredictable:** **Reproducible BiLSTM Forecasting of Incidents in the Global Terrorism Database**" details the development of a Bidirectional Long Short-Term Memory (BiLSTM) model designed to provide reliable, short-term forecasts of weekly global terrorism incidents. By leveraging nearly five decades of data from the Global Terrorism Database, the author demonstrates that this deep learning approach significantly outperforms classical statistical benchmarks by capturing complex, non-linear patterns in both the escalation and aftermath of violent events. The study identifies that long-term historical data and rolling statistical features are essential for predictive accuracy, as they allow the model to adapt to the "bursty" and non-stationary nature of geopolitical shifts [@Adegoke2025].

Building on ensemble learning techniques, the paper "**Quantitative Analysis and Prediction of Global Terrorist Attacks Based on Machine Learning**" details the development of a classification framework designed to identify and predict the specific organizations responsible for global terrorist activities. The author applies ensemble learning techniques like XGBoost and Random Forest to the GTD to analyze complex patterns in attack characteristics. The study demonstrates that these machine learning models can achieve high predictive accuracy, particularly for the most active groups, by processing variables such as location, weapon type, and target [@Pan2021].

The curators of the GTD themselves conducted their own research on using natural language processing of raw source data to populate the GTD. The article from the University of Maryland's Computer Science department, titled "**The Global Terrorism Database: Experiments in Machine-Assisted Data Collection**" tested whether natural language processing could successfully group news reports of the same event and extract specific details without constant human intervention. The findings suggest that while fully automated systems currently lack the precision of human experts, a hybrid human-AI methodology significantly reduces manual workloads and streamlines the coding process [@Colon2024].

Brandt et. al. leveraged ConfliBERT to perform natural language processing for data imputation in their research titled "**ConfliBERT: A Language Model for Political Conflict**". By integrating domain-specific expertise with natural language processing, this model excels at filtering relevant reports, identifying specific events, and performing named entity recognition to categorize key actors. Research demonstrates that ConfliBERT significantly outperforms larger, general-purpose models like Llama and Gemma in accuracy and precision while operating hundreds of times faster [@Brandt2024].

@Cai2025 developed a sophisticated AI-driven tool called FAME (Fingerprint to Article Matching for Events) to systematically track how global media outlets report on natural disasters and terrorist attacks. In their article "**Identifying and Investigating Global News Coverage of Critical Events Such as Disasters and Terrorist Attacks**" By utilizing a "fingerprint" constructed from a specific combination of event class, location, and date, the system filters millions of multilingual articles to identify those covering specific crises without the need for manual training data. This methodology reveals that news coverage is heavily biased toward wealthier nations and those with significant trade ties, often requiring much higher death tolls in poorer regions to receive equivalent attention. Ultimately, this work provides an open-source framework for computational social science to measure and understand the systemic factors that shape the global news agenda.

\clearpage

# References
